---
title: 'Assignment : Course Project 1'
author: "Manjula T"
date: "January 15, 2016 (IST)"
output: md_document
---

### 1. Code for reading in the dataset and/or processing the data

```{r, echo=TRUE}
if(!file.exists("data"))  {
	dir.create("data")
}

if(!file.exists("./data/repdatadataactivity.zip"))  {
	zipFileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip?accessType=DOWNLOAD"
	download.file(zipFileUrl, destfile="./data/repdatadataactivity.zip")
}

if(!file.exists("./data/activity.csv")) {
	## code to unzip
	unzip("./data/repdatadataactivity.zip", list=FALSE, overwrite=TRUE, exdir=".")
}

activitydata <- read.csv("./data/activity.csv")

```


### 2.Histogram of the total number of steps taken each day
```{r, echo=TRUE}
## Load plyr Package
	library(plyr)

## Group by the date and then sum the steps
  act_sum_by_date <- ddply(activitydata, "date", summarise, TotalSteps=sum(steps))

## Plot the histogram 
	hist(act_sum_by_date$TotalSteps, 
				xlab="Total Number of Steps",
				main="Histogram of Total Number of Steps taken each Day")

```

### 3.Mean and median number of steps taken each day
```{r, echo=TRUE}
## Load plyr Package
	library(plyr)

## Use Summary to Calculate the Mean and Median 
## Print the Mean and Median
  stepssummary <- summary(act_sum_by_date[,2])
  stepsmean <- stepssummary[4]
  stepsmedian <- stepssummary[3]
  print(stepsmean)
  print(stepsmedian)

```

### 4.Time series plot of the average number of steps taken
```{r, echo=TRUE}
## Load plyr, ggplot2 Package
	library(plyr)
	library(ggplot2)

## Group by the interval and then find the mean of the steps
  act_avg_by_interval <- ddply(activitydata, "interval", summarise, mean(steps,na.rm=TRUE))
  colnames(act_avg_by_interval )[2] <- "AverageSteps"
  
## Finding the range for interval
	int_range <- range(0,  act_avg_by_interval$interval)

## Plot the Graph (ggplot2)
	ThePlot <- qplot(interval, AverageSteps, data=act_avg_by_interval,
	      geom="line", method="lm",xlab="Interval", ylab="Average of Steps", 
				main="Average of Steps taken per interval", xlim=int_range)
	
	## Printing the Plot
	print(ThePlot)
	
```

### 5. The 5-minute interval that, on average, contains the maximum number of steps
```{r, echo=TRUE}
act_order <- act_avg_by_interval[order(act_avg_by_interval[,2]),]

act_max <- tail(act_order,1)
colnames(act_max )[1] <- "Interval"
colnames(act_max )[2] <- "Highest Average Step"	

print(act_max)
```

### 6. Code to describe and show a strategy for imputing missing data
### a. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r, echo=TRUE}
act_order <- act_avg_by_interval[order(act_avg_by_interval[,2]),]

thecount <- nrow(activitydata[which(is.na(activitydata$steps) == TRUE), ])

print(thecount)
```

### b. Strategy for filling in all of the missing values in the dataset. Have used the mean for that 5-minute interval, etc.
```{r, echo=TRUE}

activitydatafilled <- activitydata
thetotalCount <- nrow(activitydatafilled)

for (i in 1:thetotalCount)
{
  if (is.na(activitydatafilled[i,"steps"])==TRUE) 
	{  
     	activitydatafilled[i,"steps"]<-act_avg_by_interval[which(act_avg_by_interval$interval == activitydatafilled[i,"interval"]),2]  
  }  
}

thenewcountofNA <- nrow(activitydatafilled[which(is.na(activitydatafilled$steps) == TRUE), ])

print(thenewcountofNA)
## All NAs filled. So the new count of NA displays 0
```

### 7 (a).Histogram of the total number of steps taken each day - After imputing the missing Values
```{r, echo=TRUE}
## Load plyr Package
	library(plyr)

## Group by the date and then sum the steps
  act_sum_by_date_filled <- ddply(activitydatafilled, "date", summarise, TotalStepsFilled=sum(steps))

## Plot the histogram 
	hist(act_sum_by_date_filled$TotalStepsFilled, 
				xlab="Total Number of Steps",
				main="Histogram of Total No. of Steps taken each Day (With No missing values)")

```

### 7 (b).Mean and median number of steps taken each day - With No missing Values
```{r, echo=TRUE}
## Load plyr Package
	library(plyr)

## Use Summary to Calculate the Mean and Median 
## Print the Mean and Median
  stepssummaryfilled <- summary(act_sum_by_date_filled[,2])
  stepsmeanfilled <- stepssummaryfilled[4]
  stepsmedianfilled <- stepssummaryfilled[3]
  print(stepsmeanfilled)
  print(stepsmedianfilled)

```

### 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r, echo=TRUE}
## Load plyr Package
	library(plyr)

  thetempframe <- data.frame("daytype")
  activitydatafilled <- cbind(activitydatafilled,thetempframe)
  colnames(activitydatafilled )[4] <- "daytype"
  levels(activitydatafilled$daytype) <- c(levels(activitydatafilled$daytype), "weekday")
  levels(activitydatafilled$daytype) <- c(levels(activitydatafilled$daytype), "weekend")
  str(activitydatafilled)
  
  thetotalCount <- nrow(activitydatafilled)

  for (i in 1:thetotalCount)
  {
    
    if (weekdays(as.Date(activitydatafilled[i,"date"], "%Y-%m-%d")) == 
        "Saturday" | 
        weekdays(as.Date(activitydatafilled[i,"date"], "%Y-%m-%d")) == 
        "Sunday") 
	  {  
     	  activitydatafilled[i,"daytype"]<-"weekend"
    }
    else
    {
        activitydatafilled[i,"daytype"]<-"weekday"
    }
  }
  ## The following shows that the daytype has been filled now
  head(activitydatafilled)
  
	library(plyr)
	library(ggplot2)

  act_avg_by_interval2 <- ddply(activitydatafilled, .(interval,daytype), summarise, mean(steps,na.rm=TRUE))
  colnames(act_avg_by_interval2 )[3] <- "AverageSteps"
  
## Finding the range for interval
	int_range <- range(0,  act_avg_by_interval2$interval)
  
## Plot the Graph (ggplot2)
	ThePlot2 <- qplot(interval, AverageSteps, facets=.~daytype, data=act_avg_by_interval2,
	      geom="line", method="lm",xlab="Interval", ylab="Average of Steps", 
				main="Average of Steps taken per interval(Weekday,Weekend)", xlim=int_range)
	
	## Printing the Plot
	print(ThePlot2)

```
